The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-12 manuscript/manuscript.Rnw
discuss CGB and CGA

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CGB1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{chorionic gonadotropin}}
\label{fig:cgb_evolution_tree}
\end{figure}

@@

The growth hormone (GH) family has expanded in anthropoid primates,
resulting in approximately 5 genes in humans (GH1, GH2, CSH1, CSH2,
CSHL1). All but GH1 show substantial expression in the placenta (Table
8). There is some evidence that increased expression of these genes
may be correlated with overall birth weight of offspring in humans
\DLA{(Mannik et al., 2010)}, and it is possible that the extreme expression
levels of these genes may be due to high birth weight. Ensembl
annotations show 4 GH genes present in the chimpanzee genome, and
three of these genes show high expression in bonobo, although to a
lesser degree than in human (Table 8). Human placental GH genes are
involved in increasing blood flow, replacing maternal pituitary
expression of GH, as well as trophoblast invasion \cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth}, and it is likely that
they have the same function in the bonobo.

The human gene CGA codes for the alpha subunit of multiple hormones
including pituitary hormones (luteinizing hormone,
follicle-stimulating hormone, thyroid-stimulating hormone), as well as
numerous chorionic gonadotropins with variable beta subunits \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of the chorionic gonadotropins is to
prevent the regression of the corpus luteum \cite{Hanson.ea1971:EffectsofHCGand}.
Among the six placentas we analyzed here, CGA is only expressed in the
placentas of humans and bonobos (\DLA{Table 8}), which corresponds with the
emergence of the chorionic gonadotropins in anthropoid primates
\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly, CGA is expressed at much
higher levels in both species than the beta subunits with which its
product dimerizes (\DLA{Table 8}). CGA is one of several genes noted by
\cite{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone adaptive evolution of the
human-chimpanzee stem lineage.

Relaxin is a hormone that has multiple effects during pregnancy
including increasing maternal blood flow, widening the pubic bone and
relaxing uterine musculature in preparation for labor. In both dogs
and elephants, there is high expression of relaxin in the term
placenta, with little evidence in other species examined here (Table
8). Relaxin is known to be produced in the placenta of the pregnant
dog and is used in the diagnosis of canine pregnancy
\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof};
however, the source of relaxin produced by pregnant African elephants
remains unclear \cite{Meyer.ea2004:Speciesandfetalgender}. The high
rise in relaxin in elephants has been suggested to be a major signal
for the initiation of labor
\cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive} Here we
show that the placenta of elephants show high expression of relaxin
and is likely the predominant source of relaxin at term \DLA{(Table 8)}.

>>>>

@@

The human gene CGA codes for the alpha subunit of multiple hormones
including pituitary hormones (luteinizing hormone,
The \ac{GH} family has expanded in anthropoid primates, resulting in
approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
reliably distinguish transcripts between the two). All but
\textit{GH1} show substantial expression in the \human placenta
(\cref{fig:csh1_evolution_tree}). There is some evidence that
increased expression of these genes may be correlated with overall
birth weight of offspring in \human
\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
possible that the extreme expression levels of these genes may be due
to high birth weight. \human placental \ac{GH} genes are involved in
increasing blood flow, replacing maternal pituitary expression of
\ac{GH}, as well as trophoblast invasion
\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
and it is likely that they have the same function in \bonobo.

\begin{figure}
<<chorionic.somatomammotropin.alpha.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CGA" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{chorionic gonadotropin, alpha subunit}}
\label{fig:cga_evolution_tree}
\end{figure}

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="CGB1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{chorionic gonadotropin, beta subunit}}
\label{fig:cgb_evolution_tree}
\end{figure}

The human gene \textit{CGA} codes for the alpha subunit of multiple
hormones including pituitary hormones (luteinizing hormone,
follicle-stimulating hormone, thyroid-stimulating hormone), as well as
numerous chorionic gonadotropins with variable beta subunits
\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of
the chorionic gonadotropins is to prevent the regression of the corpus
luteum \cite{Hanson.ea1971:EffectsofHCGand}. Among the six placentas
we analyzed here, \textit{CGA} is only expressed in the placentas of
\human and \bonobo (\cref{fig:hormone_genes,fig:cga_evolution_tree}),
which corresponds with the emergence of the chorionic gonadotropins in
anthropoid primates
\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly,
\textit{CGA} is expressed at much higher levels in both species than
the beta subunits with which its product dimerizes (\cref{fig:hormone_genes,fig:cga_evolution_tree,fig:cgb_evolution_tree}).
\textit{CGA} is one of several genes noted by
\citet{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone
adaptive evolution of the \human{}-\textit{P. troglodytes} stem lineage.
including increasing maternal blood flow, widening the pubic bone and
relaxing uterine musculature in preparation for labor. In both dogs
and elephants, there is high expression of relaxin in the term
placenta, with little evidence in other species examined here (Table
8). Relaxin is known to be produced in the placenta of the pregnant
dog and is used in the diagnosis of canine pregnancy
\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof};
however, the source of relaxin produced by pregnant African elephants
remains unclear \cite{Meyer.ea2004:Speciesandfetalgender}. The high
rise in relaxin in elephants has been suggested to be a major signal
for the initiation of labor
\cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive} Here we
show that the placenta of elephants show high expression of relaxin
and is likely the predominant source of relaxin at term \DLA{(Table 8)}.
